Extracting Angular Observables without a Likelihood\\and Applications to Rare Decays

### 2014-10-21 draft.tex
[draft] some more comments and questions

For the selected class of decays, we can describe that problem as follows: The PDF $P$ has a fixed dependence on the dilepton
helicity angle $\theta_1$ and the azimuthal angle $\theta_3$. (See also \refapp{btokpill} for details
on the angular distribution.) However, at the level of decay amplitudes the dimeson system can have an arbitrarily
large total angular momentum $j$; only its third component is restricted to $m = -1,0,+1$. \fred{check}
We advocate here that is a sensible procedure to perform an expansion in terms of Legendre polynomials $p_{j}^{(|m|)}$
with respect to the remaining angle $\theta_2$. Using the observables $S_k(\vec{\nu})$, $k=1,\dots,9$, with $\vec{\nu}=(q^2, k^2)$
and as defined in \refapp{partial-waves} \fred{A lot of mixing between appendices C and E}, the expansion reads
\begin{equation}
    S_{k}(\vec{\nu},\cos\theta_2) \equiv \sum_{j} \frac{1}{n_{j,|m|}} S_{k,j}(\vec{\nu}) p_{j}^{(|m|)}(\cos\theta_2)\sin\theta_2\,,
\end{equation}
where the the normalization factor $n_{j,|m|}$ is defined in \refeq{legendre-scalar-product}. Within our
example we have (cf. also \refapp{btokpill})
\begin{equation}
    |m| = \begin{cases}
        0\,, & k = 1,2,6\,\\
        1\,, & k = 3,4,5,7,8,9\,,
    \end{cases}\,.
\end{equation}
The angular observables $S_{k,j}$ defined as coefficients of this expansion in Legendre polynomials
have the merit of a well defined total angular momentum, and thus are physically discriminable.
As a consequence of the orthogonality of the Legendre polynomials, any mismodelling (or rather, lack of modelling of higher
partial-wave observables) \emph{does not affect} the method of moments as discussed in the previous section. To wit,
ignoring further (orthogonal) terms in the PDF cannot affect the mean vector $(\dots, S_k, \dots)$ nor its variance. \fred{yes!}\\

@@

It remains to invert the relation \refeq{linear-assumption}, so that one can determine the unfolded
angular observables $\lbrace \wwhat{S_k}\rbrace$ as
\begin{equation}
    \wwhat{S_i} = \danny{N_\eps} \left(M^{-1}\right)_{ij} \est{S_j}\,.
\end{equation}
\danny{Here $N_\eps$ is a normalization factor based on the detector acceptance function $\eps$. It emerges
    since the product $P(x) \eps(x)$ is not a PDF; it is not normalized appropriately\footnote{We thank
    X.X\danny{Marcin: Please insert whatshisname} for pointing this out.}. We can determine $N_\eps$ by
    imposing
    \begin{gather}
        1 \equiv \int \rmdx{x} P(x | \lbrace \wwhat{S}_k\rbrace) = N_\eps \sum_{k,j} \left(M^{-1}\right)_{kj} \est{S_j}\int \rmdx{x} f_k(x)\,,\\
        \Rightarrow N_\eps \equiv \frac{1}{\sum_{k,j} \left(M^{-1}\right)_{kj} \est{S_j} \int \rmdx{x} f_k(x)}\,.
    \end{gather}
}

@@

\danny{In order to better understand the detector aceptance effects, we propose two steps. For the sake of
    clarity of our argument, we assume in the following that the PDF $P$ is expanded in products of
    appropriately chosen Legendre polynomials to some multi-index order $\vec{l}_P^{(0)}$. Similarly,
    moments and detector effects are nominally determined up to some order $\vec{l}_M^{(0)}$, with
    $\vec{l}_M^{(0)} \leq \vec{l}_P^{(0)}$. Here and in the following all comparisons are to be read
    for each compoenent.\\
    For the first step, we remind that the unfolding matrix $M$ depends implicitly on the physical
    PDF $P$. We therefore recommend variation of the order of the expansion of $P$ to $\vec{l}_P \geq \vec{l}_P^{(0)}$
    component-wise. This way one can estimate the mixing of PDF terms of order $\vec{l}_P$ into moments of order $\vec{l}_M^{(0)}$
    or lower.\\
    In a second step, we assume using the nominal expansion of the PDF up to order $\vec{l}_P^{(0)}$.
    With moments of order $\vec{l}_M \geq \vec{l}_M^{(0)}$ one can therefore find the impact of the
    acceptance effects on the mixing of moments of order $\vec{l}_M^{(0})$ and below
    into moments beyond the order of the PDF.
    Our recommendation is not illustrated at the hand of numerical examples, which would necessitate
    access to realistic detector models. The latter are not available to persons outside the
    respective collaborations. The proposed studies are therefore beyond the scope of this work.
}

@@

In order to study the performance of the proposed method, we simulate measurements for one uni-angular
and one tri-angular decay distribution for various numbers of events. As inputs we use the SM predictions
for the decays $B\to K\ell^+\ell^-$ and $B\to K^*(\to K\pi)\ell^+\ell^-$, respectively. We cover the
$q^2$ ranges $1\GeV^2$ to $6\GeV^2$, and $15\GeV^2$ to to the respective kinematic endpoint, with two
bin widths: $1\GeV^2$ and $0.5\GeV^2$. This is to ensure that a wide spectrum of possible values for
the angular observables.\\

Our finding can be summarized as follows:
\begin{itemize}
    \item In all studied cases we observed not a single bias in the pull distribution of the observable.
        All pull distributions can successfully be fitted to a gaussian distribution. As an example, we
        show the pull distributions for the observables $S_5$ and $S_7$ for a SM-like $B\to K^*\ell^+\ell^-$
        decays in \reffig{pulls}.
    \item We study the absolute uncertainty $\sigma_k(K)$ with respect to the angular observable $S_k$
        as a function of the number of simulated events $K$. As expected for a multivariate gaussian
        distribution, we find that the absolute uncertainty is well fitted by
        \begin{equation}
            \label{eq:unc-on-mean}
            \sigma_k(K) = \frac{\sigma_k(1)}{\sqrt{K}}
        \end{equation}
        with $\sigma_k(1) = \order{1}$, regardless of the absolute size of $S_k$. The latter can best be shown
        for the example of uncertainties of two observables. Taking again $S_5$ ($\simeq 27\%$) and $S_7$ ($\simeq 2\%$)
        for SM-like $B\to K^*\ell^+\ell^-$ decays, we show the absolute uncertainty in \reffig{errors}.
    \item For all toy studies, we produce plots of their skewness and their kurtosis as functions of
        $N$. We find in all cases that the mean skewness and kurtosis converge towards zero.
        \todo{Plot Skewness and Kurtosis for $S_5$ and $S_7$ as functions of the number of simulated events $K$.}
\end{itemize}
We find that all of the above results show consistently that the joint distrbution of the angular observables
converges towards a multivariate Gaussian distribution. We propose to describe the resulting distribution as
a multivariate Gauss when skewness and kurtosis are small enough. The precise targets for these quantities
are subject to the details of the physical analysis, and therefore outside the scope of our work.

>>>>

For the selected class of decays, we can describe that problem as follows: The PDF $P$ has a fixed dependence on the dilepton
helicity angle $\theta_1$ and the azimuthal angle $\theta_3$. (See also \refapp{btokpill} for details
on the angular distribution.) However, at the level of decay amplitudes the dimeson system can have an arbitrarily
large total angular momentum $j$; only its third component is restricted to $m = -1,0,+1$. \fred{check}
We advocate here that is a sensible procedure to perform an expansion in terms of Legendre polynomials $p_{j}^{(|m|)}$
with respect to the remaining angle $\theta_2$. Using the observables $S_k(\vec{\nu})$, $k=1,\dots,9$, with $\vec{\nu}=(q^2, k^2)$
and as defined in \refapp{partial-waves} \fred{A lot of mixing between appendices C and E}, the expansion reads
\begin{equation}
    S_{k}(\vec{\nu},\cos\theta_2) \equiv \sum_{j} \frac{1}{n_{j,|m|}} S_{k,j}(\vec{\nu}) p_{j}^{(|m|)}(\cos\theta_2)\sin\theta_2\,,
\end{equation}
\fred{since when is $S_k a fct. of \cos \theta_2$?}
where the the normalization factor $n_{j,|m|}$ is defined in \refeq{legendre-scalar-product}. Within our
example we have (cf. also \refapp{btokpill})
\begin{equation}
    |m| = \begin{cases}
        0\,, & k = 1,2,6\,\\
        1\,, & k = 3,4,5,7,8,9\,,
    \end{cases}\,.
\end{equation}
The angular observables $S_{k,j}$ defined as coefficients of this expansion in Legendre polynomials
have the merit of a well defined total angular momentum, and thus are physically discriminable.
As a consequence of the orthogonality of the Legendre polynomials, any mismodelling (or rather, lack of modelling of higher
partial-wave observables) \emph{does not affect} the method of moments as discussed in the previous section. To wit,
ignoring further (orthogonal) terms in the PDF cannot affect the mean vector $(\dots, S_k, \dots)$ nor its variance. \fred{yes!}\\

@@

It remains to invert the relation \refeq{linear-assumption}, so that one can determine the unfolded
angular observables $\lbrace \wwhat{S_k}\rbrace$ as
\begin{equation}
    \wwhat{S_i} = \danny{N_\eps} \left(M^{-1}\right)_{ij} \est{S_j}\,.
  \end{equation}
  \fred{What happens to the uncertainties? Unfolding is used in many
    places, so there are elaborate recipes. Should we assume the same
    uncertainties as before? Since it is linear, rescaling should be sufficient. Or compute covariance directly for transformed observables} \danny{Here $N_\eps$ is a normalization factor
    based on the detector acceptance function $\eps$. It emerges since
    the product $P(x) \eps(x)$ is not a PDF; it is not normalized
    appropriately\footnote{We thank X.X\danny{Marcin: Please insert
        whatshisname} for pointing this out.}. We can determine
    $N_\eps$ by imposing
    \begin{gather}
        1 \equiv \int \rmdx{x} P(x | \lbrace \wwhat{S}_k\rbrace) = N_\eps \sum_{k,j} \left(M^{-1}\right)_{kj} \est{S_j}\int \rmdx{x} f_k(x)\,,\\
        \Rightarrow N_\eps \equiv \frac{1}{\sum_{k,j} \left(M^{-1}\right)_{kj} \est{S_j} \int \rmdx{x} f_k(x)}\,.
    \end{gather}
}

@@

\danny{In order to better understand the detector aceptance effects, we propose two steps. For the sake of
    clarity of our argument, we assume in the following that the PDF $P$ is expanded in products of
    appropriately chosen Legendre polynomials to some multi-index order $\vec{l}_P^{(0)}$. Similarly,
    moments and detector effects are nominally determined up to some order $\vec{l}_M^{(0)}$, with
    $\vec{l}_M^{(0)} \leq \vec{l}_P^{(0)}$. Here and in the following all comparisons are to be read
    for each compoenent.\\
    For the first step, we remind that the unfolding matrix $M$ depends implicitly on the physical
    PDF $P$. We therefore recommend variation of the order of the expansion of $P$ to $\vec{l}_P \geq \vec{l}_P^{(0)}$
    component-wise. This way one can estimate the mixing of PDF terms of order $\vec{l}_P$ into moments of order $\vec{l}_M^{(0)}$
    or lower.\\
    In a second step, we assume using the nominal expansion of the PDF up to order $\vec{l}_P^{(0)}$.
    With moments of order $\vec{l}_M \geq \vec{l}_M^{(0)}$ one can therefore find the impact of the
    acceptance effects on the mixing of moments of order $\vec{l}_M^{(0)}$ and below
    into moments beyond the order of the PDF.
    Our recommendation is not illustrated at the hand of numerical examples, which would necessitate
    access to realistic detector models. The latter are not available to persons outside the
    respective collaborations. The proposed studies are therefore beyond the scope of this work.
}

\fred{I'm puzzled. If you can do the calculation to some order
  $\vec{l}_P$, why would you use a lower order for the actual result?
  I.e., if you have the NNLO results, why would you limit yourself to
  NLO?}

@@

In order to study the performance of the proposed method, we simulate
measurements for one uni-angular and one tri-angular decay
distribution for various numbers of events. As inputs we use the SM
predictions for the decays $B\to K\ell^+\ell^-$ and $B\to K^*(\to
K\pi)\ell^+\ell^-$, respectively. We cover the $q^2$ ranges $1 \GeV^2$
to $6 \GeV^2$, and $15 \GeV^2$ to the respective kinematic endpoint,
with two bin widths: $1\GeV^2$ and $0.5\GeV^2$. This is to ensure that
a wide spectrum of possible values for
the angular observables.\\

Our finding can be summarized as follows:
\begin{itemize}
    \item In all studied cases we observed not a single bias in the pull \fred{should define pull} distribution of the observable.
        All pull distributions can successfully be fitted to a gaussian distribution. As an example, we
        show the pull distributions for the observables $S_5$ and $S_7$ for a SM-like $B\to K^*\ell^+\ell^-$
        decays in \reffig{pulls}.
    \item We study the absolute uncertainty $\sigma_k(K)$ with respect to the angular observable $S_k$
        as a function of the number of simulated events $K$. As expected for a multivariate gaussian
        distribution, we find that the absolute uncertainty is well fitted by
        \begin{equation}
            \label{eq:unc-on-mean}
            \sigma_k(K) = \frac{\sigma_k(1)}{\sqrt{K}}
        \end{equation}
        with $\sigma_k(1) = \order{1}$, regardless of the absolute size of $S_k$. The latter can best be shown
        for the example of uncertainties of two observables. Taking again $S_5$ ($\simeq 27\%$) and $S_7$ ($\simeq 2\%$)
        for SM-like $B\to K^*\ell^+\ell^-$ decays, we show the absolute uncertainty in \reffig{errors}.
    \item For all toy studies, we produce plots of their skewness and their kurtosis as functions of
        $N$. We find in all cases that the mean skewness and kurtosis converge towards zero.
        \todo{Plot Skewness and Kurtosis for $S_5$ and $S_7$ as functions of the number of simulated events $K$.}
\end{itemize}
We find that all of the above results show consistently that the joint distrbution of the angular observables
converges towards a multivariate Gaussian distribution. We propose to describe the resulting distribution as
a multivariate Gauss when skewness and kurtosis are small enough. The precise targets for these quantities
are subject to the details of the physical analysis, and therefore outside the scope of our work.
